Union and intersection contracts: theory vs practice

### 2021-03-23 paper.tex
Fix color
Small fixes

The explorative work presented in this paper is motived by the design of the
Nickel language\cite{NickelRepo}. Nickel is a configuration programming
language, meaning that a Nickel program evaluates to a value that can be
directly serialized to JSON, YAML or alike.  Nickel aims at being lightweight,
easy to write and easy to embed in a larger system, to serve either directly as
a native language to specify configuration, or as a meta-language to generate
configuration. The focus is first and foremost practical: Nickel originated as a
rethink of the language of the Nix package manager, used to describe one of the
biggest sets of software packages ever built\cite{repology}.  Nickel targets
build systems and cloud deployment tools (Kubernetes, Terraform or NixOps) as
well. While there's no hesitation in incorporating recent research when it
serves a well-defined purpose, the design decisions are guided by the industrial
use cases.

Our aim is that writing simple Nickel should not be more complex than writing
JSON or YAML. Unlike JSON, though, Nickel anticipates large configurations by
being both programmable and typed. In the configuration setting, there is a
singular dilemma with respect to typing: since your program is run on fixed
inputs and is expected to terminate, any relevant type error will be reported
anyway.  Why bother with the complexity of a static type system? On the other
hand, more and more software systems offload complexity to configurations, such
as cloud infrastructure, build system, or operating systems, leading the
corresponding configurations to become increasingly complex, to the point where
static types are beneficial. For reusable code — that is, library functions —
static types brings all the usual, widely acknowledged benefits.

This dilemma is naturally solved by gradual typing\cite{Siek06gradualtyping}
that mixes both static typing and dynamic typing. Unlike gradually-typed
language like TypeScript, we do not seek to build a complex type system that
tries hard to make most of the idioms naturally arising in dynamically typed
code typable. Rather, we choose to provide a reasonably expressive type system
with good inference properties, that make usual generic function operating on
data easily typable, but may require resorting to untyped code for more exotic
expressions.  However, this is not a complete surrender, as Nickel provides a
complementary mechanism for more advanced data validation: contracts.

@@

Enforcing pre- and post-conditions at runtime is a widely established practice.
In their foundational paper\cite{FindlerFelleisenHOContracts}, Findler and
Fellleisen introduces a principled approach to run-time assertion checking that
nicely support higher-order functions and introduce the notion of blame, which
is fundamental for good error reporting. It turns out these contracts are
closely related to the type casts introduced by gradual typing, modulo blame:
both \cite{FindlerMultiLang} and \cite{FelleisenInterLang} see the value of
contracts as a safe interface between typed and untyped code. In
\cite{WellTypedBlamed}, the authors precisely a system integrating gradual
typing with contracts \textit{à la Findler \& Felleisen}. Nickel adopts a
similar type system, with both statically typed terms, dynamically typed terms,
and first-class contracts. In this setting, each type constructor of the static
type system - arrows, records, foralls - must have a contract counterpart that
is able to check at run-time that a value is indeed a member of the
corresponding type.

@@

>>>>

The explorative work presented in this paper is motived by the design of the
Nickel language\cite{NickelRepo}. Nickel is a configuration programming
language, meaning that a Nickel program evaluates to a value that must be
straightforwardly serializable to JSON, YAML or alike.  Nickel aims at being
lightweight, easy to write and easy to embed in a larger system, to serve either
directly as a native language to specify configuration, or as a meta-language to
generate configuration. The focus is first and foremost practical: Nickel
originated as a rethink of the language of the Nix package manager, used to
describe one of the biggest sets of software packages ever built\cite{repology}.
Nickel targets build systems and cloud deployment tools (Kubernetes, Terraform
or NixOps) as well. While there's no hesitation in incorporating recent research
when it serves a well-defined purpose, the design decisions are guided by the
industrial use cases.

\subsection{Typing}

Our aim is that writing simple Nickel should not be more complex than writing
JSON or YAML. Unlike JSON, though, Nickel anticipates large configurations by
being both programmable and typed. In the configuration setting, there is a
singular dilemma with respect to typing: since a program is run on fixed inputs
and is expected to terminate, any relevant type error should show up anyway.
Why bother with the complexity of a static type system? On the other hand, more
and more software systems offload complexity to configurations, such as the
infrastructure as code paradigm used in cloud deployment. When complexity grows,
static types becomes attractive again. For reusable code — that is, library
functions — static types brings all the usual widely acknowledged benefits.

This dilemma is naturally solved by gradual typing\cite{Siek06gradualtyping}
that mixes both static typing and dynamic typing. Unlike gradually-typed
languages like TypeScript, we do not seek to build a complex type system that
tries hard to accept most of the idioms naturally arising in dynamically typed
code. Rather, we choose to provide a reasonably expressive type system with good
inference properties, that make typing generic function operating on data easy ,
but may require to resort to untyped code for more exotic expressions.  In any
case, this is not a surrender, as Nickel provides a complementary mechanism for
more advanced data validation: contracts.

@@

@@

Enforcing pre- and post-conditions at runtime is a widely established practice.
In their foundational paper\cite{FindlerFelleisenHOContracts}, Findler and
Fellleisen introduces a principled approach to run-time assertion checking that
nicely support higher-order functions and introduce the notion of blame, which
is crucial to good error reporting. It turns out these contracts are closely
related to the type casts introduced by gradual typing, modulo blame: both
\cite{FindlerMultiLang} and \cite{FelleisenInterLang} see the value of contracts
as a safe interface between typed and untyped code. In \cite{WellTypedBlamed},
the authors precisely a system integrating gradual typing with contracts
\textit{à la Findler \& Felleisen}. Nickel adopts a similar type system, with
both statically typed terms, dynamically typed terms, and first-class contracts.
In this setting, each type constructor of the static type system - arrows,
records, foralls - must have a contract counterpart that is able to check at
run-time that a value is indeed a member of the corresponding type.
